Galois Slicing as Automatic Differentiation

### 2025-07-11 main/approx-as-tangents.tex
Tweaks.

When giving a denotational semantics for sequential programming
languages, Scott-continuous functions are too permissive. Famously,
\citet{plotkin77lcf}'s Parallel OR (\exref{parallel-or}, below) is
continuous but does not explore its input in a way consistent with a
sequential implementation. Continuous functions can explore their
input in a non-deterministic way as long as the result is
deterministic. This non-determinism results in functions whose
output cannot be assigned a unique minimal input that accounts for
it. Therefore, continuous functions are incompatible with the central
idea in \GPS that we should be able to identify a \emph{minimal} part
of the input that leads to a part of the output.

\emph{Stability} is a property that can be required of monotone functions, that was invented by \citet{berry79} in an attempt to capture sequentiality. This was unsuccessful (see the $\mathrm{gustave}$ function in \exref{parallel-or}), but we will see now how it is closely related to the problem of computing forward and backwards maps of approximations needed in \GPS. A textbook description of stable functions in the context of domain theory is given by \citet[Chapter 12]{amadio-curien}. We start with a property that is weaker than stability, but is easier to motivate in connection with derivatives of smooth functions.

>>>>

When giving a denotational semantics for sequential programming
languages, Scott-continuous functions are too permissive. Famously,
\citet{plotkin77lcf}'s Parallel OR (\exref{parallel-or}, below) is
continuous but does not explore its input in a way consistent with a
sequential implementation. Continuous functions can explore their
input in a non-deterministic way as long as the result is
deterministic. This non-determinism results in functions whose
output cannot be assigned a unique minimal input that accounts for
it. Therefore, continuous functions are in general incompatible with the central
idea in \GPS that we should be able to identify a \emph{minimal} part
of the input that leads to a part of the output.

\emph{Stability} is a property that can be required of monotone functions, that was invented by \citet{berry79} in an attempt to capture sequentiality. This was unsuccessful (see the $\mathrm{gustave}$ function in \exref{parallel-or}), but we will see now how it is closely related to the problem of computing the forward and backwards maps of approximations needed in \GPS. A textbook description of stable functions in the context of domain theory is given by \citet[Chapter 12]{amadio-curien}. We start with a property that is weaker than stability, but easier to motivate in connection with derivatives of smooth functions.

### 2025-07-11 main/language.tex
[ tex ] condensed summary in 1.3
Tweaks.

Assuming that $\cat{C}$ is bicartesian closed and has a list object (\eqnref{lists}), then we can extend an
interpretation of a signature $\Sigma$ to an interpretation of the whole language over
$\Sigma$. \figref{semantics:types} and \figref{semantics:contexts} define the interpretation of types and
contexts as objects of $\cat{C}$ respectively. Terms are interpreted as morphisms between the interpretations
of the context and type, as defined in \figref{semantics:terms}. We have used the notations $\pi_i$ for
projections, $\prodM{f}{g}$ for pairing, $\coprodM{f}{g}$ for (parameterised) copairing, $!_x$ for morphisms
to the terminal object, and $\lambda$ and $\eval$ for currying and evaluation for exponentials.

@@

\subsection{Examples}

>>>>

Assuming that $\cat{C}$ is bicartesian closed and has a list object (\eqnref{lists}), then we can extend an
interpretation of a signature $\Sigma$ to an interpretation of the whole language over
$\Sigma$. \figref{semantics:types} and \figref{semantics:contexts} define the interpretation of types and
contexts as objects of $\cat{C}$ respectively. Terms are interpreted as morphisms between the interpretations
of the context and type, as defined in \figref{semantics:terms}. We have used the notations $\pi_i$ for
projections, $\prodM{f}{g}$ for pairing, $\coprodM{f}{g}$ for (parameterised) copairing, $!_X$ for morphisms
to the terminal object, and $\lambda$ and $\eval$ for currying and evaluation for exponentials.

@@

\subsection{Examples}
\label{sec:language:examples}
